Comparing summed probability distributions of shoreline and radiocarbon dates from the Mesolithic Skagerrak coast of Norway

### 2023-09-23 analysis/paper/paper.Rmd
Cleaning repository

The most commonly applied of these is a null-hypothesis significance testing approach by means of Monte Carlo simulation, first introduced by @shennan2013 and later expanded upon by @timpson2014 and @crema2021. This works by comparing the observed RSPD with a series of simulated RSPDs, generated from a null model. These null models are typically a uniform, exponential or logistic distribution. These are chosen *a priori* with reference to common long-term population dynamics and are typically parametrised by fitting the models by means of least-squares or, given recent methodological developments [@timpson2021], maximum likelihood estimation (MLE). The result from these simulations is then used to create a 95% critical envelope representing the null model. The proportion of the observed RSPD that falls outside the simulated envelope (i.e. positive or negative divergence of the observed SPD from the fitted null model), is then used to estimate a global p-value indicating whether the null model can be rejected. In the case that it can, the portions of the observed RSPD that falls outside this envelope can subsequently be interpreted as representing potentially meaningful demographic events, relative to the null model. However, care has to be taken in how these are interpreted. First, this follows from the fact that 5% of the deviations from the critical envelope can be expected to be random, and there is no way to know which deviations this pertains to. Secondly, for example, an exponential null model fit to the data is only one of an infinite set of exponential models with different growth rates that could be used. While a model fit by means of MLE will likely have a reasonable growth rate, and by extension exclude many other exponential fits as likely to explain the data, this can be difficult to determine. Finally, the p-value only indicates whether the null model as a whole can be rejected as an explanation of the data, and it does not provide statistical justification for inferring whether local deviations themselves are meaningful demographic signals [@crema2022; @timpson2021].

@@

All analyses done in this study were performed using the R programming language [@rcoreteam]. Underlying data and programming code used for the paper is available in a version-controlled online repository at https://doi.org/10.17605/osf.io/fqwt5. This is structured as a research compendium following @marwick2018b, to allow for reproducibility of the results [see also @Marwick2017]. Analysis of the ^14^C-dates was done using the R packages *ADMUR* v1.0.3 [@timpson2021] and *rcarbon* v1.4.4 [@crema2021]. The R package *shoredate* v.1.0.2 was used for performing and handling the shoreline dating of sites [@roalkvam2023b], with re-purposed code from both *ADMUR* and *rcarbon* for subsequent Monte Carlo simulation and model comparison.

@@

Given the presence of people and corresponding signal in the SSPD before the start of the RSPD, and the suggested relevance of mobility patterns for the lacking signal in the RSPD in this earliest period, neither proxy is presumably devoid of influence from population density or mobility patterns. This is also given support by the correspondence between the logistic growth in the RSPD and the positive deviation in the SSPD in the first centuries of the 8th millennium BCE. However, properly assessing this relationship and the degree to which it has held constant throughout the period will, as mentioned, depend on independent data that track the developments of the other factors that have been suggested to be of relevance to the behaviour of the two proxies.

@@

Some further lines of inquiry could be to draw on lithic technology and tool-kit diversity as a function of population size and density [e.g. @berg-hansen2018; @collard2013a; @riede2014; @solheim2020], and mobility patterns that can potentially be inferred from the composition of lithic assemblages [e.g. @barton2014; @clark2017; @preston2018; @roalkvam2022], which is potentially also influenced by variable rates of relative sea-level change. In terms of mobility patterns, focus was directed here towards residential movement and occupational duration; but other dimensions relating for example to seasonal and territorial mobility, as reflected in the discussion of inland activity, could also prove to be important. In the same vein, environmental conditions, constraints and developments beyond relative sea-level change also represent a range of factors that can have implications for these dimensions, and will consequently be of future importance to unpack the relationship between the proxies [e.g. @hoebe2023; @jorgensen2020; @ordonez2022; @lundstrom2023]. To account for the confounding effects these might have, a more principled and comprehensive evaluation of how investigatory practices might impact the SPDs would also be beneficial. While certainly also pertinent for the RSPD, this has generally been considered more extensively for radiocarbon data than for shoreline dated sites. Furthermore, aligning the complex web of causes and effects that these different lines of evidence potentially represent will also benefit from a more direct modelling of the development in the SSPD, rather than simply basing the analysis on deviations from a rejected null model [cf. @timpson2021]. Finally, mapping and contrasting these developments across a larger region will not only offer a larger sample size, which is especially limited for the ^14^C-data employed here, but could also provide important contextualisation for the observed development [see e.g. @lundstrom2021; @manninen2023; @solheim2018] and potentially reduce the impact of idiosyncratic investigations that might impact the present SPDs inordinately [see e.g. @crema2022; @hoebe2023; @rick1987].

>>>>

The most commonly applied of these is a null-hypothesis significance testing approach by means of Monte Carlo simulation, first introduced by @shennan2013 and later expanded upon by @timpson2014 and @crema2021. This works by comparing the observed RSPD with a series of simulated RSPDs, generated from a null model. These null models are typically a uniform, exponential or logistic distribution. These are generally chosen *a priori* with reference to common long-term population dynamics and are typically parametrised by fitting the models by means of least-squares or, given recent methodological developments [@timpson2021], maximum likelihood estimation (MLE). The result from these simulations is then used to create a 95% critical envelope representing the null model. The proportion of the observed RSPD that falls outside the simulated envelope (i.e. positive or negative divergence of the observed SPD from the fitted null model), is then used to estimate a global p-value indicating whether the null model can be rejected. In the case that it can, the portions of the observed RSPD that falls outside this envelope can subsequently be interpreted as representing potentially meaningful demographic events, relative to the null model. However, care has to be taken in how these are interpreted. First, this follows from the fact that 5% of the deviations from the critical envelope can be expected to be random, and there is no way to know which deviations this pertains to. Secondly, for example, an exponential null model fit to the data is only one of an infinite set of exponential models with different growth rates that could be used. While a model fit by means of MLE will likely have a reasonable growth rate, and by extension exclude many other exponential fits as likely to explain the data, this can be difficult to determine. Finally, the p-value only indicates whether the null model as a whole can be rejected as an explanation of the data, and it does not provide statistical justification for inferring whether local deviations themselves are meaningful demographic signals [@crema2022; @timpson2021].

@@

All analyses done in this study were performed using the R programming language [@rcoreteam]. Underlying data and programming code used for the paper is available in an online repository at https://doi.org/10.17605/osf.io/fqwt5. This is structured as a research compendium following @marwick2018b, to allow for reproducibility of the results [see also @Marwick2017]. Analysis of the ^14^C-dates was done using the R packages *ADMUR* v1.0.3 [@timpson2021] and *rcarbon* v1.4.4 [@crema2021]. The R package *shoredate* v1.0.2 was used for performing and handling the shoreline dating of sites [@roalkvam2023b], with re-purposed code from both *ADMUR* and *rcarbon* for subsequent Monte Carlo simulation and model comparison.

@@

Given the presence of people and corresponding signal in the SSPD before the start of the RSPD, and the suggested relevance of mobility patterns for the lacking signal in the RSPD in this earliest period, neither proxy is presumably devoid of influence from population density or mobility patterns. This is also given support by the correspondence between the logistic growth in the RSPD and the positive deviation in the SSPD in the first centuries of the 8th millennium BCE. However, properly assessing this relationship and the degree to which it has held constant throughout the period will depend on independent data that track the developments of the other factors that have been suggested to be of relevance to the behaviour of the two proxies.

@@

Some further lines of inquiry could be to draw on lithic technology and tool-kit diversity as a function of population size and density [e.g. @berg-hansen2018; @collard2013a; @riede2014; @solheim2020], and mobility patterns that can potentially be inferred from the composition of lithic assemblages [e.g. @barton2014; @clark2017; @preston2018; @roalkvam2022], which is potentially also influenced by variable rates of relative sea-level change. In terms of mobility patterns, focus was directed here towards residential movement and occupational duration; but other dimensions relating for example to seasonal and territorial mobility, as reflected in the discussion of inland activity, could also prove to be important. In the same vein, environmental conditions, constraints and developments beyond relative sea-level change also represent a range of factors that can have implications for these dimensions, and will consequently be of future importance to unpack the relationship between the proxies [e.g. @hoebe2023; @manninen2018; @jorgensen2020; @ordonez2022; @lundstrom2023]. To account for the confounding effects these might have, a more principled and comprehensive evaluation of how investigatory practices might impact the SPDs would also be beneficial. While certainly also pertinent for the RSPD, this has generally been considered more extensively for radiocarbon data than for shoreline dated sites. Furthermore, aligning the complex web of causes and effects that these different lines of evidence potentially represent will also benefit from a more direct modelling of the development in the SSPD, rather than simply basing the analysis on deviations from a rejected null model [cf. @timpson2021]. Finally, mapping and contrasting these developments across a larger region will not only offer a larger sample size, which is especially limited for the ^14^C-data employed here, but could also provide important contextualisation for the observed development [see e.g. @lundstrom2021; @manninen2023; @solheim2018] and potentially reduce the impact of idiosyncratic investigations that might impact the present SPDs inordinately [see e.g. @crema2022; @hoebe2023; @rick1987].
